Supplementary Information

### 2020-09-13 doc/manuscript/MEE_manuscript.Rmd
incorporate JM changes

[Kristina J. Anderson-Teixeira^1,2^*](0000-0001-8461-9713),
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Camille Piponiot^1,2^,
Ross Alexander,
Craig D. Allen,
Raquel Alfaro-Sánchez,
Tala Awada,
Jennifer Baltzar,
Patrick Baker,
Sarayudh Bunyavejchewin,
Paolo Cherubini,
Justin Cooper,
Stuart Davies^2^,
Ryan Helcoski^1^,
[Jakub Kašpar^cz^](0000-0003-1780-6310),
Kamil Kral,
James Lutz,
Ellis Q. Margolis,
Justin Maxwell,
Sean McMahon^2^,
Helene Muller-Landau^2^,
Sabrina Russo,
[Pavel Šamonil^cz^](0000-0002-7722-8797),
Anastasia Sniderhan,
Alan J. Tepley^1,at^,
[Ivana Vašíčková^cz^](0000-0002-6070-5956), 
Mart Vlam,
Peter Zuidema

@@

@@

**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.** 
Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--most commonly temperature and precipitation--over the current year and, typically, previous year (e.g., DENDRO_REFS). 
While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS) and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g.,TREE_LEVEL_REFS; @banbury_morgan_global_nodate].
Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.

@@

**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_nodate], yet traditional methods do not characterize its effect or its potential interactions with climate.** 
For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
While convenient for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015; @mcgregor_tree_nodate], and smaller trees can be more sensitive to temperature (*Rollinson et al. in review*).
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
To use tree-rings to predict tree growth and $ANPP_{woody}$, we need models that include $DBH$ and its potential climate interactions.

@@

**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).**
For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS). 
Specifically, we used a generalized additive model (GAM) to fit a spline to individual growth records ($\Delta r$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $\Delta r$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*). (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
For each group of candidate climate variables, we move forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

The top model was used to visualize the effects of each variables while the others are kept constant at the median in the various figures of this manuscript. 
For these visualizing tools, we also fetched the t-test's p-value for each term of the model and represented the significance using different line types. 
Model predictions represented were back-transformed to show the effect on the original scale of $\Delta r$, $BAI$, or $\Delta AGB$.

*Data and code availability*

*Data*, code, and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
Data for # of the sites are archived in the The DendroEcological Network (DEN) database [@rayback_dendroecological_2020].

@@

![**Figure 3 | Species-level responses of $\Delta r$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (THIS FIGURE WILL PROBABLY CHANGED INTO A 4COL X 5 ROW MATIX, DROPPING SPECIES KEY, WHICH CAN BE SEEN IN OTHER FIGURES)](tables_figures/pre_temp_groups.png)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, DASHED lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

All this points to the potential importance of considering DBH in climate reconstructions.
Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
*(Indeed, this is probably one factor contributing to the non-stationarity of climate signals in tree-ring records [@wilmking_global_2020].)*
*(	Interactive effects of DBH and climate may suggest significant impact of different sampling strategy on results per individual site. -- e.g., stronger drought signal if just large trees are sampled. This is consistent with the dendro tradition of sampling canopy trees for strongest drought signals; however, our results suggest that temperature signals may be stronger in young trees.  )*
*(height vs exposure/ social status of trees)*

@@

>>>>

[Kristina J. Anderson-Teixeira^1,2^*](0000-0001-8461-9713),
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Camille Piponiot^1,2^,
Ross Alexander,
Craig D. Allen,
Raquel Alfaro-Sánchez,
Tala Awada,
Jennifer Baltzar,
Patrick Baker,
Sarayudh Bunyavejchewin,
Paolo Cherubini,
Justin Cooper,
Stuart Davies^2^,
Ryan Helcoski^1^,
[Jakub Kašpar^cz^](0000-0003-1780-6310),
Kamil Kral,
James Lutz,
Ellis Q. Margolis,
Justin Maxwell^jm^,
Sean McMahon^2^,
Helene Muller-Landau^2^,
Sabrina Russo,
[Pavel Šamonil^cz^](0000-0002-7722-8797),
Anastasia Sniderhan,
Alan J. Tepley^1,at^,
[Ivana Vašíčková^cz^](0000-0002-6070-5956), 
Mart Vlam,
Peter Zuidema

@@

jm. Department of Geography, Indiana University, Bloomington, Indiana, USA

@@

**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.** 
Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--most commonly temperature, precipitation, or a moisture index (PDSI or SPEI)--over the current year and, typically, previous year [e.g., @fritts_tree_1976]. 
While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions [e.g., @pederson_epic_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2017; @banbury_morgan_global_nodate].
Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.

@@

**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_nodate], yet traditional methods do not characterize its effect or its potential interactions with climate.** 
For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
While convenient for identifying climate signals [(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015; @mcgregor_tree_nodate], and smaller trees can be more sensitive to temperature (*Rollinson et al. in review*).
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
To use tree-rings to predict tree growth and $ANPP_{woody}$, we need models that include $DBH$ and its potential climate interactions.

@@

**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).**
For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers [@fritts_tree_1976]. 
Specifically, we used a generalized additive model (GAM) to fit a spline to individual growth records ($\Delta r$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $\Delta r$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*). (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
For each group of candidate climate variables, we move forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.

@@

@@

![**Figure 3 | Species-level responses of $\Delta r$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. For each relationship shown, other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (THIS FIGURE WILL PROBABLY CHANGED INTO A 4COL X 5 ROW MATIX, DROPPING SPECIES KEY, WHICH CAN BE SEEN IN OTHER FIGURES)](tables_figures/pre_temp_groups.png)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model. Other terms in the model are held constant at their median. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), DASHED lines when only one term is significant, and dotted lines when neither is significant.  Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

All this points to the potential importance of considering DBH in climate reconstructions *[but these reconstructions are often selecting trees at much bigger dbhs that what are at these sites.]*
Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
*(Indeed, this is probably one factor contributing to the non-stationarity of climate signals in tree-ring records [@wilmking_global_2020].)*
*(	Interactive effects of DBH and climate may suggest significant impact of different sampling strategy on results per individual site. -- e.g., stronger drought signal if just large trees are sampled. This is consistent with the dendro tradition of sampling canopy trees for strongest drought signals; however, our results suggest that temperature signals may be stronger in young trees.  )*
*(height vs exposure/ social status of trees)*

@@

### Data availability
```{r eval=FALSE, echo=FALSE}
# To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.
```

*To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.*

Code and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
Data for # of the sites are archived in the The DendroEcological Network (DEN) database [@rayback_dendroecological_2020].
